### 2020-06-20 chapters/background.tex
...
EMAs draft

\subsection{Streaming Windows}

@@

In the query above, the
\texttt{TUMBLE(timestamp, Time.minutes(5), rack\_id)} statement groups incoming data using time-based tumbling windows of 5 minutes duration, using the \texttt{timestamp} field to evict events not in this time interval. The resulting list of rows has \texttt{rack\_id} as its key and as corresponding value a list of temperatures, from the past 5 minutes. Finally, for each rack and its list of temperatures, we compute an \texttt{AVG} aggregation, resulting in the average temperature per rack in the past 5 minutes.

@@

Subtract-On-Evict (SOE) on the other hand, is a SWAG algorithm that incrementally computes the aggregation value. For each element inserted or evicted from the window, the aggregation value is updated, in contrast to RFS which would recompute the entire aggregation. SOE relies on expiring the evicted elements from the aggregation state. Hence, SOE only works for \textit{invertible} aggregations, that is, a function $\ominus$ exists such that $(x \oplus y) \ominus y = x$, for all $x$ and $y$. To illustrate, consider the computation of a sum aggregation over a sliding window. The total sum of the window will be the aggregation state or value. When inserting an element, the updated total window sum equals the current total sum plus ($\oplus$) the new element. The sum aggregation is considered \textit{invertible} because, when evicting an element, the new aggregation state is the total sum computed thus far minus ($\ominus$) the evicted element. With window size of \textit{n}, SOE has \textit{O(n)} space complexity and \textit{O(1)} time complexity. However, aggregations are not always invertible thus making this algorithm very restrictive and not always applicable. Algorithm \ref{pseudo:soe} shows pseudocode for SOE.

Subtract-On-Evict is an improvement over the Recalculate-From-Scratch algorithm based on the idea of aggregation invertibility. The question then becomes if there are other aggregation properties besides invertibility that must be taken into account when analyzing general stream processing algorithms.

@@

@@

\begin{table}[!htb]
\begin{center}
     \begin{tabular}{p{2cm}|p{1.5cm}p{1.5cm}p{1.5cm}p{1.5cm}p{1.6cm}|}
                 & \rotatebox{35}{\textbf{invertible combine}} & \rotatebox{35}{\textbf{associative combine}} & \rotatebox{35}{\textbf{commutative combine}} & \rotatebox{35}{\textbf{size-preserving combine}} & \rotatebox{35}{\textbf{unary query}}
                 \\
                 \hline
    \textbf{sum-like}     & \checkmark & \checkmark               & \checkmark               & \checkmark                   & \checkmark       \\
    \textbf{collect-like} & \checkmark              & \checkmark               & $\times$               & $\times$                   & ?           \\
    \textbf{median-like}  & \checkmark              & \checkmark               & \checkmark               & $\times$                   & ?           \\
    \textbf{max-like}     & $\times$              & \checkmark               & ?                   & \checkmark                   & \checkmark       \\
    \textbf{sketch-like}  & $\times$              & \checkmark               & \checkmark               & \checkmark                   & $\times$      
    \end{tabular}
    \caption[Aggregation properties]{Aggregation properties. Checkmarks (\checkmark), crosses ($\times$), and question marks (?) indicate a property is true for all, false for all, or false for some of a given group of like operations, respectively}
  \label{tbl:aggregations-properties}
    \end{center}
\end{table}

@@

\textcolor{red}{TODO: explain EMAs, give mathematical proofs on error bounds, emulated window size based on nhalf}

>>>>

\subsection{Sliding Windows}

@@

In the query above, the
\texttt{TUMBLE(timestamp, Time.minutes(5), rack\_id)} statement groups incoming data using time-based tumbling windows of 5 minutes duration, using the \texttt{timestamp} field to evict events not in this time interval. The resulting list of rows has \texttt{rack\_id} as its key and as corresponding value a list of temperatures, from the past 5 minutes. Finally, for each rack and its list of temperatures, we compute an \texttt{AVG} aggregation, resulting in the average temperature per rack in the past 5 minutes. Note that this is very similar to any aggregation made in a standard Database Management System (DBMS). The main difference is that in standard SQL the \texttt{GROUP BY} clause includes only the columns on which the data will be grouped while in streaming SQL the \texttt{GROUP BY} also specifies the type of window used (\textit{e.g.} in this case, a tumbling window).

@@

Subtract-On-Evict (SOE) on the other hand, is a SWAG algorithm that incrementally computes the aggregation value. For each element inserted or evicted from the window, the aggregation value is updated without recomputing the entire aggregation like in RFS. SOE relies on expiring the evicted elements from the aggregation state. Hence, SOE only works for \textit{invertible} aggregations, that is, a function $\ominus$ exists such that $(x \oplus y) \ominus y = x$, for all $x$ and $y$. To illustrate the differences between SOE and RFS, consider the computation of a sum aggregation over a sliding window. The total sum of the window will be the aggregation state or value. When inserting an element, the updated total window sum equals the current total sum plus ($\oplus$) the new element. The sum aggregation is considered \textit{invertible} because, when evicting an element, the new aggregation state is the total sum computed thus far minus ($\ominus$) the evicted element. With window size of \textit{n}, SOE has \textit{O(n)} space complexity and \textit{O(1)} time complexity. However, aggregations are not always invertible thus making this algorithm very restrictive and not always applicable. Algorithm \ref{pseudo:soe} shows pseudocode for SOE.

@@

On the previous Section, we presented Subtract-On-Evict as an improvement over Recalculate-From-Scratch based on the idea of aggregation invertibility. The question then becomes if there are other aggregation properties besides invertibility that must be taken into account when analyzing general stream processing algorithms.

@@

\begin{table}[!htb]
\begin{center}
     \begin{tabular}{p{2cm}|p{1.5cm}p{1.5cm}p{1.5cm}p{1.5cm}p{1.6cm}|}
                 & \rotatebox{35}{\textbf{invertible combine}} & \rotatebox{35}{\textbf{associative combine}} & \rotatebox{35}{\textbf{commutative combine}} & \rotatebox{35}{\textbf{size-preserving combine}} & \rotatebox{35}{\textbf{unary query}}
                 \\
                 \hline
    \textbf{sum-like}     & \checkmark & \checkmark               & \checkmark               & \checkmark                   & \checkmark       \\
    \textbf{collect-like} & \checkmark              & \checkmark               & $\times$               & $\times$                   & ?           \\
    \textbf{median-like}  & \checkmark              & \checkmark               & \checkmark               & $\times$                   & ?           \\
    \textbf{max-like}     & $\times$              & \checkmark               & ?                   & \checkmark                   & \checkmark       \\
    \textbf{sketch-like}  & $\times$              & \checkmark               & \checkmark               & \checkmark                   & $\times$      
    \end{tabular}

    \end{center}
    \caption[Aggregation properties]{Aggregation properties. Checkmarks (\checkmark), crosses ($\times$), and question marks (?) indicate a property is true for all, false for all, or false for some of a given group of like operations, respectively}
  \label{tbl:aggregations-properties}
\end{table}

@@

We talked about sliding window aggregations and algorithms to compute them and observe that the contents of the sliding window must be stored in memory so that when a new element is inserted we can evict the oldest one and discount its effect from the aggregation value. In this Section we analyze Exponential Moving Averages (EMAs). EMAs process each event only upon insertion. Since there is no eviction operation to perform, we do not need to store the whole window of data.

Exponential Moving Averages (also known as Exponential Weighted Moving Averages) \cite{EMA-Everett2011} \cite{EMA-Hunter} \cite{EMA-MarcusB} are applied to a sequence of elements to compute a weighted average, using different weights for each element. These weights decrease exponentially with the age of the observations. In other words, EMAs give higher weights to more recent data points.

\textcolor{red}{Sliding window Weights image}
\textcolor{red}{EMA Weights image}

EMAs are recursive: the EMA value at timestamp \textit{t} depends only on the EMA value at timestamp \textit{t-1}. This recursive property allows us to store only one EMA value and continuously update it for each new value val\textsubscript{t}, based on the following recursion formula:

\begin{definition}
EMA\textsubscript{t} = EMA\textsubscript{t-1} * $\lambda$ + val\textsubscript{t}

\label{def:ema}
\end{definition}

with $\lambda$ as a constant factor applied to the previous EMA value. This $\lambda$ constant (0 < $\lambda \leq$ 1) is known as the smoothing coefficient. The smoothing or decay factor $\lambda$ is given by:

\[ \lambda = 2^{- \dfrac{1}{\tau}} \]

where $\tau$ is the half-life of the EMA. The half-life $\tau$ controls how fast older events are forgotten as the smoothing factor depends on it. The half-life is the amount of time that it takes for an element to see its contribution or weight reduced to half. For instance, with $\tau = 1s$ we have $\lambda = 0.5$, which means for each 1s step we reduce the weights applied to each event by half.

% EMA-count example (later build histogram concept on multiple EMA-counts)

### 2020-06-20 chapters/my-work.tex
EMAs draft

In our lightweight real-time solution we need constant time complexity to process events one by one in real-time. Additionally, because our system is not mission-critical, we want a solution as lightweight as possible, which means something that grows linearly with window size is not a good fit. Therefore, most of the outlier detection methods presented become unfit for our use due to high memory consumption and time complexity. Some of the methods shown also suffer from what we call \textit{"memory loss"}: they lose track of the reference window or normality period. In other words, the algorithms are online learners, meaning they adapt to the underlying statistics of the data stream. While in some scenarios this is a much-needed feature, in our use-case it is not. How clear the alerts and insights retrieved from the methods are is also an important property. Some outlier detection methods used Machine Learning (ML) techniques to produce alerts, which makes it harder to explain to a system administrator why an alert was raised and what changed.

@@

Our goal is to detect data pattern shifts in an online fashion. Such shifts will be measured between a reference or normality period versus the observed reality during the streaming phase. Note that we do not need to directly compare the entire window contents of the reference period with the contents of the sliding window. The latter would require storing all window contents hence making the memory consumption grow linearly with sliding window and/or reference period size. Instead, we want to use an aggregation that does not grow linearly in memory regarding window size and that can be incrementally maintained during the streaming phase. The resulting reference and streaming aggregations should be comparable with the intent of raising divergence alerts. In short, we want a reference aggregation snapshot to compare with a streaming aggregation that must be updated in real-time and constant in memory usage.

To that end, we devise a two-phase method. The first phase relies on a batch analysis that builds our aggregated reference snapshot to be later on compared with the sliding one. The second phase focus on efficiently updating the sliding window aggregation snapshot and occasionally perform a similarity test between the static reference aggregation and the sliding one, alerting for changes if need be. The proposed method currently works only for numerical features, that is, only for event fields or variables that are of numeric type, but can be extended to other types, as we will discuss.

\subsection{Stream monitoring goals}
\textcolor{red}{TODO: add some pseudocode}

Before diving into the batch and streaming algorithms we must further define our goal. In this Section, we address the question: \textit{"What is our goal during the streaming or monitoring phase?"}

A data stream is an infinite stream of events. Each event is timestamped and contains multiple fields. In the context of this Thesis, each of the event's fields are also named features or time-dependent variables. They can be thought of as columns in a traditional relational database table. Our goal is to produce alerts of the following form:

\[\textit{timestamp: [(x1, 0.23), (x2, 0.51), (x3, 0.55)}]\]

This indicates the \textit{timestamp} for the given alert, which features diverge (in this case \textit{x1, x2, x3}) and a value representative of how much each feature diverges (\textit{e.g.} a float value between 0 and 1 that works as a confidence value).

An alert should be raised if a feature distribution changes considerably. This immediately poses challenges such as \textit{(a)} finding a suitable aggregation that encodes each feature's distribution and allows for a change detection test and \textit{(b)} defining a deviation threshold $\alpha$.

\subsection*{Approximated Histogram Aggregation} \label{sec:ema-hist}
\textcolor{red}{TODO: add some pseudocode}

We chose to encode a feature distribution as a histogram. A histogram is an aggregation that represents the distribution of a numerical feature or a set of categories. However, we must recall that we need to update this histogram in constant time in a sliding window fashion, implying the eviction of old elements and insertion of new ones must be done with \textit{O(1)} time complexity.

An exact histogram aggregation must keep the entire sliding window in memory...

\textcolor{red}{TODO: review this section after EMAs is written}

\subsection*{Defining the deviation threshold}

Assume we use a distance function that takes two histograms --- our reference aggregation histogram and our sliding window aggregation histogram --- and outputs a distance value. How do we threshold this value? For instance, consider the following left and right-skewed distance distributions represented by Figures \ref{fig:skewed-left-distro} and \ref{fig:skewed-right-distro}, respectively. Note that these histograms do not represent the reference and/or sliding window aggregation histograms. Here we make the assumption we have a histogram that encodes the distribution of distance values measured between the reference aggregation and the sliding window aggregation.

@@

Hard-coding a threshold value to be used for all features is cumbersome, impractical and error-prone. The distance threshold must be computed based on the distribution of distances. How do we get to know the distribution of such distances? That is one of the goals for the batch phase.

@@

The goal of the batch phase will be to build our reference snapshot and get to know the distribution of "distances" between the reference snapshot and the sliding window one. In other words, we obtain a reference aggregation histogram that we, later on, use to measure its distance to the sliding window aggregation histogram, producing a distance value. We compare the measured distance with its distribution to find out how probable it is and alert in a very unlikely scenario. We do this for each feature.

@@

For each sample, and for each feature, we compute the approximated histogram using the bins computed for the reference histogram (Figure \ref{fig:ref-hist-bins}.

@@

Recall how an approximated histogram is built as explained in Section \ref{sec:ema-hist}. We will process each sample event by event and apply a discount factor per event, thus building an approximated histogram aggregation using the reference bins, for each feature of that sample. We compute multiple sample histograms to encode distributions over smaller time periods that may exist within the large reference period. We use an approximated histogram to mimic the target approximated histogram we will incrementally maintain in the streaming environment.

@@

For each feature, we end up with \textit{S} distance values. These values are distance measurements between random samples of data and the reference period, and the histograms for each sample were built exactly like they will be maintained in streaming for better fidelity. Hence, we claim we now know the distribution of expected distance values. Given a new distance value measured between another window and the reference window, we can compute its probability and produce alerts if below a certain threshold.

>>>>

In our lightweight real-time solution we need constant time complexity to process events one by one in real-time. Additionally, because our system is not mission-critical, we want a solution as lightweight as possible, which means something that grows linearly with window size is not a good fit. Therefore, most of the outlier detection methods presented become unfit for our use due to high memory consumption and time complexity. Some of the methods shown also suffer from what we call \textit{"memory loss"}: they lose track of the reference window or normality period. In other words, the algorithms are online learners, meaning they adapt to the underlying statistics of the data stream. While in some scenarios this is a much-needed feature, in our use-case it is not. How clear the alerts and insights retrieved from the methods are, is also an important property. Some outlier detection methods used Machine Learning (ML) techniques to produce alerts, which makes it harder to explain to a system administrator why an alert was raised and what changed.

@@

A data stream is a continuous collection of events. Each event is timestamped and contains multiple fields. In the context of this Thesis, each of the event's fields are time-dependent variables named features. They can be thought of as columns in a traditional relational database table. Our goal is to detect data pattern shifts for each feature in an online fashion. Such shifts will be measured between a reference period versus the observed reality during the streaming phase. Figure \ref{fig:timelines} shows the separated reference and streaming timelines.

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.4]{figures/timeperiods.png}
      \caption[]{Reference and Streaming timelines}
      \label{fig:timelines}
    \end{center}
\end{figure}

We do not need to directly compare the entire window contents of the reference and sliding window periods. The latter would require storing all window contents hence making the memory consumption grow linearly with sliding window and/or reference period size. Instead, we want to use a  \textit{(a)} constant in memory aggregation that can be \textit{(b)} incrementally maintained during the streaming phase. Additionally, we want a streaming aggregation that \textit{(c)} reliably encodes the distribution of each feature.

Histograms are a representation of the distribution of numerical or categorical data thus being ideal to encode the distribution of feature values, satisfying \textit{(c)}. Histograms require the storage of a fixed-size list of bins and a corresponding fixed-size list of counts per bin for ever growing volumes of data, satisfying property \textit{(a)}. Adding or removing an element to the histogram aggregation is tantamount to determining the correct bin and adding or subtracting one, respectively, thus satisfying property \textit{(b)}. However, computing an histogram aggregation over a sliding window of size $W$ implies storing all $W$ elements in memory to know the element to remove after a new one is inserted. This violates our constant in memory property \textit{(a)} which is why we chose to use the Exponential Moving Averages (EMAs) analyzed in Section \ref{sec:emas}. EMA-like histograms computed over sliding windows do not require the storage of window contents. An EMA-like histogram is \textit{(a)} constant in memory, \textit{(b)} maintained in real-time and \textit{(c)} encodes the distribution of each feature values. EMA-like histograms are further explained in Section \ref{sec:ema-hist}.

Since the reference window is static, \textit{i.e.}, it is not a sliding window, we can use exact histogram aggregations. Hence, for each feature, we build an exact reference histogram aggregation. On the other hand, in streaming, we use one EMA-like approximate histogram aggregation per feature (Figure \ref{fig:timelines-hists}).

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.4]{figures/timelines-hists.png}
      \caption[]{Reference and Streaming timelines and corresponding histogram aggregations}
      \label{fig:timelines-hists}
    \end{center}
\end{figure}

To compare each feature's reference and sliding window histogram aggregation we need a measure of divergence. We chose to use the Jensen–Shannon Divergence (JSD) \cite{JSD} but other statistical tests are possible such as Kolmogorov-Smirnov or Wasserstein. \textcolor{red}{defer the why to later section} Given a divergence value for the two aggregations (reference and sliding window ones) we need a measure of normal statistical fluctuations. Since we assume a reference period as the normal state we can find out the normal distribution for the divergence by sampling smaller windows of data from the reference period in a batch analysis. This allows us to define a threshold based on percentiles, which gives an estimate for the probability of a certain value occurring, rather than using a constant hand picked threshold for the divergence value. \textcolor{red}{defer details to later section}

We devise a two-phase method. The first phase relies on a batch analysis that, for each feature, builds a reference histogram and computes the distribution of divergence values (details presented in Section ??). The second phase focus on efficiently updating the sliding window histogram for each feature and periodically performs a divergence test between the static reference aggregation and the sliding one, alerting for changes if need be. The proposed method currently works only for numerical features, that is, only for event fields or variables that are of numeric type, but can be extended to other types.

\iffalse
\subsection*{Defining the deviation threshold}
Assume we use a distance function that takes two histograms --- our reference aggregation histogram and our sliding window aggregation histogram --- and outputs a distance value. How do we threshold this value? For instance, consider the following left and right-skewed distance distributions represented by Figures \ref{fig:skewed-left-distro} and \ref{fig:skewed-right-distro}, respectively. Note that these histograms do not represent the reference and/or sliding window aggregation histograms. Here we make the assumption we have a histogram that encodes the distribution of distance values measured between the reference aggregation and the sliding window aggregation.

@@

Hard-coding a threshold value to be used for all features is cumbersome, impractical and error-prone. The distance threshold must be computed based on the distribution of distances. How do we get to know the distribution of such distances? That is one of the goals of the batch phase.
\fi

\subsection{Exponential Moving Average based histogram} \label{sec:ema-hist}
The sliding window histogram we use is an Exponential Moving Average (EMA) based histogram. An EMA-like histogram is essentially a collection of EMA-counts as described in Section \ref{sec:emas}.

.....

@@

The goal of the batch phase will be to build our reference aggregation for each feature and get to know the distribution of divergence or distance values between the reference and the sliding window aggregations. In other words, we:

\begin{enumerate}
    \item obtain a reference histogram for each feature using all data in the reference period

    \item then compute the distribution of distances between the reference histogram and the sliding window histogram by computing the distance (or divergence value) between the reference histogram and various samples of sliding window histograms
\end{enumerate}

This distribution of values can then later be compared to a given value of divergence for a single sample to find out how probable that value is to occur (if it is very low an alert may be raised).

@@

For each sample, and for each feature, we compute the approximated histogram using the bins computed for the reference histogram (Figure \ref{fig:ref-hist-bins}). We also add two extra bins that will cover the region to the left and to the right of our histogram, respectively. In other words, we add a bin that covers the region between $-\infty$ and our first bin and another that covers values from the last bin to $+\infty$. This way, observed values outside of the reference period will be placed in these special bins, either left or right.

@@

Each of the histograms is an EMA-like histogram. An EMA-like histogram is essentially a collection of EMA-counts, as described in Section \ref{sec:ema-hist}. Each sample is processed event by event and a discount factor is applied per event, thus building an approximated histogram aggregation, using the reference bins, for each feature of that sample. We compute multiple sample histograms to encode distributions over smaller time periods that may exist within the large reference period. We use an approximated histogram to mimic the target approximated histogram we will incrementally maintain in the streaming environment.

@@

For each feature, we end up with \textit{S} distance values. These values are distance measurements between random samples of data and the reference period. Hence, we claim we now know the distribution of expected distance values. Given a new distance value measured between another random window and the reference window, we can compute its probability and produce alerts if below a certain threshold. Note that the histograms for each sample are EMA-like histograms (as detailed in Section \ref{sec:ema-hist}) just like the sliding window histogram. This ensures a certain degree of fidelity in the test we make in streaming because we measure the divergence between a random sample of data (our streaming histogram or any of the sample's histogram) and the reference one.

### 2020-06-20 chapters/sota.tex
EMAs draft

A time-series is a collection of data points indexed by time. Accordingly, the authors define a time-series as \textit{"observations that have been recorded in an orderly fashion and which are correlated in time"}. Regarding the concept of \textit{outlier}, the authors state that \textit{"outliers can be thought of as observations that do not follow the expected behavior"}. This definition is similar to the concept of outlier provided by Hawkins in 1980 where he states that an outlier is \textit{“an observation which deviates so much from other observations as to arouse suspicions that it was
generated by a different mechanism"} \cite{Hawkins-Outliers}. Furthermore, Blázquez-García et. al allude to the fact that outliers are often referred to in the literature as \textit{"anomalies, discordant observations, discords, exceptions, aberrations, surprises, peculiarities or contaminants"} which for the purpose of their review and this Thesis all refer to outliers. In summary, outlier detection refers to the process of finding anomalies in data. An anomaly (or outlier) relates to a point in time where the behavior of the system does not match the expected one. Note that under this definition the presence of outliers or anomalies does not necessarily imply a problem.

@@

The \textit{input data} axis refers to the type of time-series the method uses as input. According to this criterion, an outlier detection method is said to work with a univariate time-series or a multivariate one. A univariate time-series refers to a time-series consisting of a list of values generated by a single time-dependent variable throughout time. To quote the authors, a univariate time-series is \textit{"an ordered set of real-valued observations, where each observation is recorded at a specific time"}. On the other hand, a multivariate time-series has more than one time-dependent variable. Most importantly, each variable is not only time-dependent but may also have dependencies on other variables. With that in mind, the authors define a multivariate time-series as \textit{"an ordered set of k-dimensional vectors, each of which is recorded at a specific time"}. Each \textit{k-dimensional} vector consists of \textit{k}  time-dependent variables, possibly correlated.

The \textit{nature of the method} axis refers to the type of analysis performed by the outlier detection method. An outlier detection method may perform a univariate or multivariate analysis. A univariate method works only with a single time-dependent variable whereas a multivariate one simultaneously works with more than one variable. An outlier detection method can take as input a multivariate time-series but perform a univariate analysis, one per time-series of the multivariate input. On the other hand, to perform multivariate analysis, the input data type must be multivariate as well.

@@

The authors split point outlier detection methods into uni and multivariate, as seen in Figure \ref{fig:outlier-taxonomy}. Furthermore, they split univariate point outlier detection methods into three categories: model-based, density-based and histogram-based \ref{fig:point-outlier-taxonomy}.

@@

The univariate point outlier detection methods based on models are the most common. For each timestamp of the time-series, the models compute the expected value for the time-dependent variable and compare it with the observed value. If the difference between the expected value and the observed one is above a certain threshold the data point at the current timestamp is considered an outlier. Mathematically, given the expected value $e\textsubscript{t}$, the actually observed value $x\textsubscript{t}$ and the user-defined threshold $\alpha$, $x\textsubscript{t}$ is considered an outlier at timestamp \textit{t} if and only if
\begin{equation*}
    |x\textsubscript{t}-e\textsubscript{t}|>\alpha
\end{equation*}
Additionally, the authors split the model-based category into estimation and prediction models. The main difference is that estimation model-based methods use past and future data while prediction model-based methods use only past data and infer current and future values. Because of this, prediction model-based outlier detection methods can be used in real-time to process streaming time-series, while estimation based methods can't.

@@

Ideally we want to maintain for each data point \textit{p\textsubscript{i}}, the number of neighbors it has, rather than the actual list of neighbors. The main advantage of this approach is the reduced memory footprint since only a count is stored rather than a list of pointers to all neighbors. The problem is that when a data point \textit{p\textsubscript{i}} expires, the count of each of \textit{p\textsubscript{i}}'s neighbors must be decremented by one but we are unable to reach \textit{p\textsubscript{i}}'s neighbors without a list of neighbor pointers. The solution to this problem is exploiting the previously presented \textit{predictability} property. The key idea is to predict the expiration of any data point \textit{p\textsubscript{i}} and pre-handle the impact it has on \textit{p\textsubscript{i}}'s neighbors. The authors introduce the concept of \textit{"lifetime neighbor counts"} or \textit{lt\_cnt}. The \textit{lt\_cnt} of a data point \textit{p\textsubscript{i}} corresponds to a sequence of \textit{predicted neighbor counts} --- \textit{i.e.} the number of \textit{predicted neighbors} \textit{p\textsubscript{i}} has in any future window where he exists. More precisely, each entry on \textit{pi.lt\_cnt} records the number of \textit{p\textsubscript{i}}'s neighbors that are known to survive in the corresponding future window. The authors provide the following example: consider a window \textit{W\textsubscript{i}} and a data point \textit{p\textsubscript{i}} with three neighbors, \textit{p\textsubscript{1}}, \textit{p\textsubscript{2}} and \textit{p\textsubscript{3}}. Applying the \textit{predictability} property makes it possible to compute the lifespan of all points. Assume \textit{p\textsubscript{1}} expires after \textit{W\textsubscript{i}}, \textit{p\textsubscript{2}} and \textit{p\textsubscript{3}} expire after \textit{W\textsubscript{i+1}} and \textit{p\textsubscript{i}} expires after \textit{W\textsubscript{i+2}}. This implies that at \textit{W\textsubscript{i}}, 
\[ \textit{p\textsubscript{i}.lt\_cnt} = \{\textit{W\textsubscript{i}}: 3, \textit{W\textsubscript{i+1}}: 2, \textit{W\textsubscript{i+2}}: 0\} \]
The lifetime neighbor counts (\textit{lt\_cnt}) carry enough information to compute density-based point outliers: for each data point \textit{p\textsubscript{i}} check if the first or current element of \textit{p\textsubscript{i}.lt\_cnt} is less than a user-defined threshold $\alpha$ to decide whether a data point is an outlier or not.

>>>>

A time-series is a collection of data points sorted by time. Regarding the concept of \textit{outlier}, the authors state that \textit{"outliers can be thought of as observations that do not follow the expected behavior"}. This definition is similar to the concept of outlier provided by Hawkins in 1980 where he states that an outlier is \textit{“an observation which deviates so much from other observations as to arouse suspicions that it was
generated by a different mechanism"} \cite{Hawkins-Outliers}. Furthermore, Blázquez-García et. al allude to the fact that outliers are often referred to in the literature as \textit{"anomalies, discordant observations, discords, exceptions, aberrations, surprises, peculiarities or contaminants"} which for the purpose of their review and this Thesis all refer to outliers. In summary, outlier detection refers to the process of finding anomalies in data. An anomaly (or outlier) relates to a point in time where the behavior of the system does not match the expected one. Note that under this definition the presence of outliers or anomalies does not necessarily imply a problem.

@@

The \textit{input data} axis refers to the type of time-series the method uses as input. According to this criterion, an outlier detection method is said to work with a univariate time-series or a multivariate one. A univariate time-series refers to a time-series consisting of a list of values generated by a single time-dependent variable throughout time. On the other hand, a multivariate time-series has more than one time-dependent variable. Most importantly, each variable is not only time-dependent but may also have dependencies on other variables.

The \textit{nature of the method} axis refers to the type of analysis performed by the outlier detection method. An outlier detection method may perform a univariate or multivariate analysis. A multivariate analysis considers all variables in the time-series simultaneously while univariate analysis looks at each variable independently. An outlier detection method can take as input a multivariate time-series but perform a univariate analysis, one per time-series of the multivariate input. On the other hand, to perform multivariate analysis, the input data type must be multivariate as well.

@@

Univariate point outlier detection methods are further broken down into three categories: model-based, density-based and histogram-based (Figure \ref{fig:point-outlier-taxonomy}).

@@

The univariate point outlier detection methods based on models are the most common. For each timestamp of the time-series, the models compute the expected value for the time-dependent variable and compare it with the observed value. If the difference between the expected value and the observed one is above a certain threshold the data point at the current timestamp is considered an outlier. Mathematically, given the expected value $e\textsubscript{t}$, the actually observed value $x\textsubscript{t}$ and the user-defined threshold $\alpha$, $x\textsubscript{t}$ is considered an outlier at timestamp \textit{t} if and only if
\begin{equation*}
    |x\textsubscript{t}-e\textsubscript{t}|>\alpha
\end{equation*}
Model-based point outlier detection methods can be one of two types: estimation or prediction models. The main difference is that estimation model-based methods use past and future data while prediction model-based methods use only past data and infer current and future values. Because of this, prediction model-based outlier detection methods can be used in real-time to process streaming time-series, while estimation based methods can not.

@@

The authors want to maintain for each data point \textit{p\textsubscript{i}} the number of neighbors it has, rather than the actual list of neighbors. The main advantage of this approach is the reduced memory footprint since only a count is stored rather than a list of pointers to all neighbors. The problem is that when a data point \textit{p\textsubscript{i}} expires, the count of each of \textit{p\textsubscript{i}}'s neighbors must be decremented by one but we are unable to reach \textit{p\textsubscript{i}}'s neighbors without a list of neighbor pointers. The solution to this problem is exploiting the previously presented \textit{predictability} property. The key idea is to predict the expiration of any data point \textit{p\textsubscript{i}} and pre-handle the impact it has on \textit{p\textsubscript{i}}'s neighbors. The authors introduce the concept of \textit{"lifetime neighbor counts"} or \textit{lt\_cnt}. The \textit{lt\_cnt} of a data point \textit{p\textsubscript{i}} corresponds to a sequence of \textit{predicted neighbor counts} --- \textit{i.e.} the number of \textit{predicted neighbors} \textit{p\textsubscript{i}} has in any future window where he exists. More precisely, each entry on \textit{pi.lt\_cnt} records the number of \textit{p\textsubscript{i}}'s neighbors that are known to survive in the corresponding future window. The authors provide the following example: consider a window \textit{W\textsubscript{i}} and a data point \textit{p\textsubscript{i}} with three neighbors, \textit{p\textsubscript{1}}, \textit{p\textsubscript{2}} and \textit{p\textsubscript{3}}. Applying the \textit{predictability} property makes it possible to compute the lifespan of all points. Assume \textit{p\textsubscript{1}} expires after \textit{W\textsubscript{i}}, \textit{p\textsubscript{2}} and \textit{p\textsubscript{3}} expire after \textit{W\textsubscript{i+1}} and \textit{p\textsubscript{i}} expires after \textit{W\textsubscript{i+2}}. This implies that at \textit{W\textsubscript{i}}, 
\[ \textit{p\textsubscript{i}.lt\_cnt} = \{\textit{W\textsubscript{i}}: 3, \textit{W\textsubscript{i+1}}: 2, \textit{W\textsubscript{i+2}}: 0\} \]
The lifetime neighbor counts (\textit{lt\_cnt}) carry enough information to compute density-based point outliers: for each data point \textit{p\textsubscript{i}} check if the first or current element of \textit{p\textsubscript{i}.lt\_cnt} is less than a user-defined threshold $\alpha$ to decide whether a data point is an outlier or not.

### 2020-06-20 chapters/validation.tex
EMAs draft

\section{Datasets description}
\subsection{Synthetic datasets}
\subsection{Financial fraud datasets}

\section{Tests description}
\subsection{Single feature}
\subsection{Multi feature}

\section{Tests results}
\subsection{Single feature}
\subsection{Multi feature}

>>>>

\section{Datasets}

We generated multiple synthetic datasets before moving on to real data. Furthermore, we first performed single feature analysis before moving to a multi-feature scenario. The problem of multiple testing (Section \ref{sec:multi-test}) only arises in the latter case so we only apply the Holm-Bonferroni correction (Section \ref{sec:holmbonferroni}) for that one.

\begin{table}[!htb]
    \begin{center}
        \begin{tabular}{|c|c|c|}
        \hline
        \textbf{Dataset} & \textbf{Features} & \textbf{Events} \\ \hline
        R1               & 1                 & 5.000.000       \\ \hline
        T1               & 1                 & 5.000.000       \\ \hline

        \end{tabular}
    \end{center}
    \caption{Summary statistics for each dataset}
    \label{tbl:summary-statistics}
\end{table}

\subsection{Synthetic datasets}
Dataset R1 had a single feature that followed a normal distribution with mean $\mu=10$ and $\sigma=2$ for the entirety of the dataset. Figure \ref{fig:timeseries-r1} shows the time-series of the single feature in dataset R1.

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.6]{figures/01-reference.png}
      \caption[]{Time-series for dataset R1}
      \label{fig:timeseries-r1}
    \end{center}
\end{figure}

Dataset T1 also had 5 million events with a single feature. For the first half, \textit{i.e.}, for the first 2.5 million events, the feature followed a normal distribution with mean $\mu=10$ and $\sigma=2$. For the other half, we changed the generating distribution to be a continuous uniform one with $a=100$ and $b=200$. Figure \ref{fig:timeseries-t1} shows the time-series of the single feature in dataset T1 where we see the abrupt change in feature values.

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.6]{figures/01-target.png}
      \caption[]{Time-series for dataset T1}
      \label{fig:timeseries-t1}
    \end{center}
\end{figure}

\subsection{Financial fraud datasets}

\section{Experiments}
\textcolor{red}{one subsection per experiment with i) purpose, ii) execution, iii) results obtained and what thye mean}

\subsection{Single-feature analysis with varying sample sizes}
In this first experiment, we used dataset R1 as the reference period and dataset T1 as the target or streaming period. We used 100 bins for the equal-height reference histogram and then added the two special bins mentioned in Section \ref{sec:sampling-batch}. We changed the half-life of each EMA-like sample histogram (recall Section \ref{sec:ema-hist}), effectively changing its size. We tested samples with \textit{(a)} 2.500, \textit{(b)} 250.000 and \textit{(c)} 1.000.000 events.

\textcolor{red}{We use tukey threshold, how do we introduce this, ahhhhhhhhhhhh}
